---
title: "CodeBook.md"
author: "Ana Jaime"
date: "November 14, 2017"

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r directory}
setwd("~/COURSERA/JHU-Data Science Specialization/GIT BASH/GettingAndCleaningDataProject")
```



                                    DATA DICTIONARY
                                    
    1. ORIGINAL DATASET: Human Activity Recognition Using Smartphones Dataset
        1.1. Data description
        1.2. Original variables
        1.3. Original datasets
        1.4. Final notes
    2. TIDY DATASET: Merged information with activity and variables labels
        2.1. Transformations made to the original dataset
        2.2. Variables
        2.3. Units
        2.4. Final tidy_data
    3. INDEPENDENT TIDY DATASET: Average of each measurement variable by subject and activity
        3.1. Transformations made to the tidy_data set
        3.2. Variables
        3.3. Units
        3.4. Final ind_tidy_data
    4. FOR MORE INFORMATION



## 1. ORIGINAL DATASET: Human Activity Recognition Using Smartphones Dataset

### 1.1. Data description

The following is the original description of the "Human Activity Recognition Using Smartphones Dataset" features.

**Source: *"features_info.txt"* file from**
https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip

The features selected for this database come from the accelerometer and gyroscope 3-axial raw signals tAcc-XYZ and tGyro-XYZ. These time domain signals (prefix 't' to denote time) were captured at a constant rate of 50 Hz. Then they were filtered using a median filter and a 3rd order low pass Butterworth filter with a corner frequency of 20 Hz to remove noise. Similarly, the acceleration signal was then separated into body and gravity acceleration signals (tBodyAcc-XYZ and tGravityAcc-XYZ) using another low pass Butterworth filter with a corner frequency of 0.3 Hz. 

Subsequently, the body linear acceleration and angular velocity were derived in time to obtain Jerk signals (tBodyAccJerk-XYZ and tBodyGyroJerk-XYZ). Also the magnitude of these three-dimensional signals were calculated using the Euclidean norm (tBodyAccMag, tGravityAccMag, tBodyAccJerkMag, tBodyGyroMag, tBodyGyroJerkMag). 

Finally a Fast Fourier Transform (FFT) was applied to some of these signals producing fBodyAcc-XYZ, fBodyAccJerk-XYZ, fBodyGyro-XYZ, fBodyAccJerkMag, fBodyGyroMag, fBodyGyroJerkMag. (Note the 'f' to indicate frequency domain signals). 

These signals were used to estimate variables of the feature vector for each pattern:  
'-XYZ' is used to denote 3-axial signals in the X, Y and Z directions.

* tBodyAcc-XYZ
* tGravityAcc-XYZ
* tBodyAccJerk-XYZ
* tBodyGyro-XYZ
* tBodyGyroJerk-XYZ
* tBodyAccMag
* tGravityAccMag
* tBodyAccJerkMag
* tBodyGyroMag
* tBodyGyroJerkMag
* fBodyAcc-XYZ
* fBodyAccJerk-XYZ
* fBodyGyro-XYZ
* fBodyAccMag
* fBodyAccJerkMag
* fBodyGyroMag
* fBodyGyroJerkMag


The set of variables that were estimated from these signals are: 

* mean(): Mean value
* std(): Standard deviation
* mad(): Median absolute deviation 
* max(): Largest value in array
* min(): Smallest value in array
* sma(): Signal magnitude area
* energy(): Energy measure. Sum of the squares divided by the number of values. 
* iqr(): Interquartile range 
* entropy(): Signal entropy
* arCoeff(): Autorregresion coefficients with Burg order equal to 4
* correlation(): correlation coefficient between two signals
* maxInds(): index of the frequency component with largest magnitude
* meanFreq(): Weighted average of the frequency components to obtain a mean frequency
* skewness(): skewness of the frequency domain signal 
* kurtosis(): kurtosis of the frequency domain signal 
* bandsEnergy(): Energy of a frequency interval within the 64 bins of the FFT of each window.
* angle(): Angle between to vectors. 


Additional vectors obtained by averaging the signals in a signal window sample. These are used on the angle() variable:

* gravityMean
* tBodyAccMean
* tBodyAccJerkMean
* tBodyGyroMean
* tBodyGyroJerkMean


### 1.2. Original variables

The complete list of variables of each feature vector is available in *"features.txt"*, as shown here:

**Source: *"features.txt"* file from**
https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip

```{r features}

setwd("~/COURSERA/JHU-Data Science Specialization/GIT BASH/GettingAndCleaningDataProject/UCI HAR Dataset")

features<-read.table("features.txt")

setwd("~/COURSERA/JHU-Data Science Specialization/GIT BASH/GettingAndCleaningDataProject")

print("RUN: features")
features
```


### 1.3. Original datasets

The original datasets are stored in the following objects : 'subject_test','x_test','y_test','subject_train','x_train' and 'y_train';  which are created by the code contained in the *"run_analysis.R"* script:


```{r original_data}

setwd("~/COURSERA/JHU-Data Science Specialization/GIT BASH/GettingAndCleaningDataProject/UCI HAR Dataset/test")
subject_test<-read.table("subject_test.txt")
x_test<-read.table("X_test.txt")
y_test<-read.table("y_test.txt")

setwd("~/COURSERA/JHU-Data Science Specialization/GIT BASH/GettingAndCleaningDataProject/UCI HAR Dataset/train")
subject_train<-read.table("subject_train.txt")
x_train<-read.table("X_train.txt")
y_train<-read.table("y_train.txt")

setwd("~/COURSERA/JHU-Data Science Specialization/GIT BASH/GettingAndCleaningDataProject/UCI HAR Dataset")
activity_names<-read.table("activity_labels.txt")

setwd("~/COURSERA/JHU-Data Science Specialization/GIT BASH/GettingAndCleaningDataProject")

print("RUN: str(subject_test)")
str(subject_test)
print("RUN: str(x_test)")
str(x_test)
print("RUN: str(y_test)")
str(y_test)
print("RUN: str(subject_train)")
str(subject_train)
print("RUN: str(x_train)")
str(x_train)
print("RUN: str(y_train)")
str(y_train)
print("RUN: str(activity_names)")
str(activity_names)

```



### 1.4. Final notes

The following is the final information included in the *"README.txt"* file:

**Source: *"README.txt"* file from**
https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip

The experiments have been carried out with a group of 30 volunteers within an age bracket of 19-48 years. Each person performed six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. Using its embedded accelerometer and gyroscope, we captured 3-axial linear acceleration and 3-axial angular velocity at a constant rate of 50Hz. The experiments have been video-recorded to label the data manually. The obtained dataset has been randomly partitioned into two sets, where 70% of the volunteers was selected for generating the training data and 30% the test data. 

The sensor signals (accelerometer and gyroscope) were pre-processed by applying noise filters and then sampled in fixed-width sliding windows of 2.56 sec and 50% overlap (128 readings/window). The sensor acceleration signal, which has gravitational and body motion components, was separated using a Butterworth low-pass filter into body acceleration and gravity. The gravitational force is assumed to have only low frequency components, therefore a filter with 0.3 Hz cutoff frequency was used. From each window, a vector of features was obtained by calculating variables from the time and frequency domain. See 'features_info.txt' for more details. 

For each record it is provided:

* Triaxial acceleration from the accelerometer (total acceleration) and the estimated body acceleration.
* Triaxial Angular velocity from the gyroscope. 
* A 561-feature vector with time and frequency domain variables. 
* Its activity label. 
* An identifier of the subject who carried out the experiment.

Notes: 

* Features are normalized and bounded within [-1,1].
* Each feature vector is a row on the text file.

For more information about this dataset contact: activityrecognition@smartlab.ws

License:

Use of this dataset in publications must be acknowledged by referencing the following publication [1] 

[1] Davide Anguita, Alessandro Ghio, Luca Oneto, Xavier Parra and Jorge L. Reyes-Ortiz. Human Activity Recognition on Smartphones using a Multiclass Hardware-Friendly Support Vector Machine. International Workshop of Ambient Assisted Living (IWAAL 2012). Vitoria-Gasteiz, Spain. Dec 2012

This dataset is distributed AS-IS and no responsibility implied or explicit can be addressed to the authors or their institutions for its use or misuse. Any commercial use is prohibited.

Jorge L. Reyes-Ortiz, Alessandro Ghio, Luca Oneto, Davide Anguita. November 2012.




## 2. TIDY DATASET: Merged information with activity and variables labels


### 2.1. Transformations made to the original dataset

The original datasets are **merged** at a first moment into the 'test' and 'train' subsets. 


The 'test' database results from merging (cbind) the objects mentioned in section 1.3.: 'subject_test', 'x_test' and 'y_test'


```{r test}

names(subject_test)<-c("subject")
names(x_test)<-features[,2]
names(y_test)<-c("activity_labels")
test<-cbind(subject_test,x_test)
test<-cbind(test,y_test)

print("RUN: str(test)")
str(test)

```


and it contains 2947 obs. of these 563 variables:

* A **'subject'** column/variable for which each row identifies the subject who performed the activity for the test sample. Its range is from 1 to 30. 
* An **'activity_labels'** column/variable with the activity performed by each subject for the test sample. Its range is from 1 to 6.
* 561 columns/variables, each one corresponding to one feature/measurement variable. **These variables are labeled** based on the original features list from the **ORIGINAL DATASET**. This modification aims to make variable names more descriptive in comparison to the original ones.


The 'train' database results from merging (cbind) the objects mentioned in section 1.3.: 'subject_train', 'x_train' and 'y_train'


```{r train}

names(subject_train)<-c("subject")
names(x_train)<-features[,2]
names(y_train)<-c("activity_labels")
train<-cbind(subject_train,x_train)
train<-cbind(train,y_train)

print("RUN: str(train)")
str(train)

```


and it contains 7352 obs. of these 563 variables:

* A **'subject'** column/variable for which each row identifies the subject who performed the activity for the test sample. Its range is from 1 to 30. 
* An **'activity_labels'** column/variable with the activity performed by each subject for the test sample. Its range is from 1 to 6.
* 561 columns/variables, each one corresponding to one feature/measurement variable. **These variables are labeled** based on the original features list from the **ORIGINAL DATASET**. This modification aims to make variable names more descriptive in comparison to the original ones.


Finally, the **'complete_data'** object merges (rbind) 'test' and 'train' sample sets into one complete data set containing in total 10299 obs. of 563 variables.

```{r complete}

complete_data<-rbind(train,test)

print("RUN: str(complete_data)")
str(complete_data)

```


Once the **'complete_data'** database is obtained, those **variables that estimate mean values or standard deviation values are preserved**, while the rest of the measurement variables are eliminated. This transformation aims to facilitate the manipulation of the data that it is important to the analysis. In order to do this, variables labels/names are evaluated to identify those that contain the words 'mean()' or 'std()' on it. The new dataset is stored in an object called **'tidy_data**

After this, the **'activity_labels' variable is tranformed** into a new variable named **'activity'**. This transformation consists on replacing each number representing an activity, by the activity name itself in order to make the variable content more descriptive. In order to do this, it is used the classification given in the file **"activity_labels.txt"**

**Source: *"activity_labels.txt"* file from**
https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip

where V1 refers to the level (1 to 6), and V2 refers to the activity that correspond to each level:

```{r labels}

setwd("~/COURSERA/JHU-Data Science Specialization/GIT BASH/GettingAndCleaningDataProject/UCI HAR Dataset")
activity_names<-read.table("activity_labels.txt")

print("RUN: activity_names")
activity_names

```

The 'activity_labels' variable is then removed, while the new **'activity'** variable is tranformed into a **factor**. On the other hand, the **'subject'** variable is set to be a **factor** even though there is no hierarchy.

From the merging process, variables were labeled. Further information can be found in section 2.2. The final dataset is stored in the object called **'tidy_data'**.


### 2.2. Variables

The labeling provided in the feature vector from the file *"features.txt"*

**Source: *"features.txt"* file from**
https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip

is used in order to label the dataset **measurement variables (columns 2 to 67)**. This labeling provides appropiate descriptive names for each measurement variable, and for this reason it is not modified. In other words, these variable names are found to be sufficiently descriptive and to bring comprenhensive information about each measurement variable. Specifically,each component of the name provides specific and accurate information:

* 1st component: a letter indicates the variable domain (t sets for time and f sets for frequency).
* 2nd component: gives information about the motion components (Body sets for body motion and Gravity sets for gravitational motion). 
* 3rd component: tells whether the sensor signals were captured by the accelerometer or by the gyroscope (Acc sets for accelerometer and Gyro sets for Gyroscope). 
* 4th component (present just in some cases): tells whether Jerk signals or magnitude were calculated (Jerk sets for Jerk signals and Mag sets for magnitude).
* 5th component: gives information about the set of variables that were estimated from the signals captured (mean() sets for mean value and std() sets for standard deviation). 
* 6th component (present just in some cases): tells the direction of the signal captured (X sets for x-axis direction, Y sets for y-axis direction and Z sets for z-axis direction)


The variable that identifies the **subject who performed the activity (column 1)** was labeled as "subject" which is considered to be a sufficiently descriptive name for this variable in particular.

The variable that identifies the **activity performed by each subject (column 68)** was labeled as "activity" which is considered to be a sufficiently descriptive name for this variable in particular.

These are the final **variables of the TIDY DATASET** stored in the **'tidy_data'** object:

```{r tidy}

mean_positions<-grep("mean()",names(complete_data),value=TRUE)
a<-grep("meanFreq",mean_positions,invert = TRUE)
mean_positions<-mean_positions[a]
mean_positions<-which(names(complete_data)%in%mean_positions)
std_positions<-grep("std()",names(complete_data))
b<-which(names(complete_data)=="subject")
c<-which(names(complete_data)=="activity_labels")
final_columns<-c(b,c,mean_positions,std_positions)
tidy_data<-complete_data[,final_columns]


activity_names$V2<-as.character(activity_names$V2)

for(i in 1:length(tidy_data$activity_labels)){
  for(j in 1:6){
if(tidy_data$activity_labels[i]==j){
 tidy_data$activity[i]<-activity_names[j,2] 
    }
  }
}


tidy_data<-tidy_data[,-which(names(tidy_data)=="activity_labels")]
tidy_data$subject<-as.factor(tidy_data$subject)
tidy_data$activity<-as.factor(tidy_data$activity)

print("RUN: names(tidy_data)")
names(tidy_data)

```



### 2.3. Units

'subject' is a factor variable that identifies the subject who perfomed the activity. The ID number ranges from 1 to 30.

```{r subject}

print("RUN: str(tidy_data$subject)")
str(tidy_data$subject)

```


'activity' is a factor variable that identifies the activity performed by each subject. It has 6 levels, 1 for each posible activity:


```{r activity}

print("RUN: str(tidy_data$activity)")
str(tidy_data$activity)

print("RUN: levels(tidy_data$activity)")
levels(tidy_data$activity)

```

The original units domain of the measurements variables were time (t-) **s** and frequency (f-) **Hz**. Nevertheless, since the **data was normalized** and bounded within [-1,1] from the **ORIGINAL DATASET**, the units of measurement were eliminated at a first moment enabling easily future comparation. In this way, all the measurement variables are numeric:


```{r units}

print("RUN: str(tidy_data[2:67])")
str(tidy_data[2:67])

```



### 2.4. Final tidy_data

Final **'tidy_data'** base contains 10299 obs. of 68 variables (2 factor variables and 66 numeric variables):


```{r tidy_show}

print("RUN: str(tidy_data)")
str(tidy_data)

```



## 3. INDEPENDENT TIDY DATASET: Average of each measurement variable by subject and activity


### 3.1. Transformations made to the tidy_data set

The **'tidy_data'** set is **collapsed** into a new independent database that summarizes the average of each measurement variable by activity and subject called **'ind_tidy_data'**. 

Also, variable names are modified: the word *average:* is added at the beginning of each variable name (for measurement variables). In this way, names are updated in order to give accurate information about which kind of calculation is now shown by the summarized variables. 


### 3.2. Variables

Besides the little transformation made to variable names (the word *average:* is added at the beginning of each variable name), the labeling structure of measurements variables is not modified. Further information about variable names' components and structure can be found in section 2.2.

These are the final **variables of the INDEPENDENT TIDY DATASET** stored in the **'ind_tidy_data'** object:

```{r ind}

library(plyr)
ind_tidy_data<-ddply(tidy_data,.(activity,subject),numcolwise(mean))

for(i in 3:length(names(ind_tidy_data))){
names(ind_tidy_data)[i]<-paste("average:",names(ind_tidy_data)[i])
}

print("RUN: names(ind_tidy_data)")
names(ind_tidy_data)

```



### 3.3. Units

'subject' is a factor variable that identifies the subject who perfomed the activity. The ID number ranges from 1 to 30.

```{r ind_subject}

print("RUN: str(ind_tidy_data$subject)")
str(ind_tidy_data$subject)

```


'activity' is a factor variable that identifies the activity performed by each subject. It has 6 levels, 1 for each posible activity:


```{r ind_activity}

print("RUN: str(ind_tidy_data$activity)")
str(ind_tidy_data$activity)

print("RUN: levels(ind_tidy_data$activity)")
levels(ind_tidy_data$activity)

```

The original units domain of the measurements variables were time (t-) **s** and frequency (f-) **Hz**. Nevertheless, since the **data was normalized** and bounded within [-1,1] from the **ORIGINAL DATASET**, the units of measurement were eliminated at a first moment enabling easily future comparation. In this way, all the measurement variables are numeric:


```{r ind_units}

print("RUN: str(ind_tidy_data[2:67])")
str(ind_tidy_data[2:67])

```



### 3.4. Final ind_tidy_data

Final **'ind_tidy_data'** base contains 180 obs. (30 subjects, each one performing 6 different activities) of 68 variables (2 factor variables and 66 numeric variables):


```{r ind_tidy_show}

print("RUN: str(ind_tidy_data)")
str(ind_tidy_data)

```



## 4. FOR MORE INFORMATION

For more information about 'tidy_data' and 'ind_tidy_data' sets contact: https://github.com/AnaMJaimeR/





